---
title: "16S Analysis"
output: html_document
date: "2025-06-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Processing for Alpha Diversity, Maaslin3, Beta Diversity

Requires"asv_table_cp_20240625.rds" and "amylase_metadata_list.csv"
Install and load necessary libraries

```{r cars, incldue=FALSE}

if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("biobakery/maaslin3")
for (lib in c('maaslin3', 'dplyr', 'ggplot2', 'knitr', 'kableExtra')) {
  suppressPackageStartupMessages(require(lib, character.only = TRUE))
}
BiocManager::install("decontam")
install.packages("readr")
library(readr)
library(maaslin3)
library(decontam)
library(ggpubr)
library(dplyr)
library(rstatix)

```

## Load and Decontam Data and Metadata Files

Load in Amplicon Data
```{r, include=FALSE}
df <- readRDS("asv_table_cp_20240625.rds")
df <- asv_table_cp_20240625$Amplicon

```

## Define Decomtam Function

```{r}
decontaminate_phyloseq <- function(ps4decontam){
  # Get blanks
  ps4decontam <- subset_samples(ps4decontam, sample.type != "zymo-control")
  print(ps4decontam)
  sample_data(ps4decontam)$is.neg <- sample_data(ps4decontam)$Sample_or_Control ==
    "Control-Sample"
  # Get dataframe of contaminants
  contamdf.prev05 <- isContaminant(
    ps4decontam,
    method="prevalence",
    neg="is.neg",
    threshold=0.5
  )
  print(table(contamdf.prev05$contaminant))
  print(filter(contamdf.prev05, contaminant == TRUE) %>% rownames())
  # Prune contaminant taxa at the threshold .05 (> prevalence in control=contaminant)
  final_biom <- prune_taxa(!contamdf.prev05$contaminant, ps4decontam)
  print(ntaxa(final_biom) / ntaxa(ps4decontam))
  # final_biom <- subset_samples(final_biom, sample.type != "zymo-control")
  return(final_biom)
}
#Decontam
df <- decontaminate_phyloseq(df)

# Filter out unassigned/eukaryote/chloroplast/mitochondria
df <- phyloseq::subset_taxa(df, !Kingdom %in% c("Unassigned", "d__Eukaryota"))
df <- phyloseq::subset_taxa(df, !Genus %in% c("Chloroplast", "Mitochondria"))

```


## Load metadata table (assigns each sample to high or low-intermediate group for amylase)
## If you cannot locate file path, go to file.choose() and copy path into the read.csv() below

```{r}
amylase_metadata <- read.csv("Amylase_metadata_list_ambiguityremoved.csv")

#Ensure Sample Names for Data/Metadata Overlap
samples_in_df <- sample_names(df)
samples_in_meta <- amylase_metadata[,1]

# Determine common samples between data and metadata
common_samples <- intersect(samples_in_df,samples_in_meta) 

# Subset the data file to only include shared samples
df_pruned <- subset_samples(df, sample_names(df) %in% common_samples)

```


## Agglomerate Counts at Genus Level and Assign Taxonomy

```{r}
#Agglomerate at the Genus level. This sums counts for all OTUs that share the same Genus.
df_genus <- tax_glom(df_pruned, taxrank = "Genus")

# Extract the genus level OTU table.
otu_genus <- otu_table(df_genus)

# Ensure that the OTU table orientation is correct (samples as rows)
if(taxa_are_rows(otu_genus)) {
  genus_counts <- as.data.frame(t(otu_genus))
} else {
  genus_counts <- as.data.frame(otu_genus)
}

# Update the column names using the Genus names from the taxonomy table.
tax_genus <- tax_table(df_genus)[, "Genus"]
colnames(genus_counts) <- as.character(tax_genus)

# Convert to character vector
genus_names <- as.character(tax_genus)

# Identify duplicated "uncultured" entries and make them unique
uncultured_indices <- which(genus_names == "uncultured")
if (length(uncultured_indices) > 0) {
  genus_names[uncultured_indices] <- paste0("uncultured_", seq_along(uncultured_indices))
}

# Assign updated names to the count table
colnames(genus_counts) <- genus_names

# Inspect the first few rows of the family-level count table.
head(genus_counts)

# Combine duplicate genus columns by summing their counts across rows
genus_counts <- as.data.frame(
  t(apply(genus_counts, 1, function(x) tapply(x, names(x), sum)))
)

# Inspect the first few rows of the genus-level count table.
head(genus_counts)

```

## Agglomerate Counts at Family Level and Assign Taxonomy

```{r, include=FALSE}
#Agglomerate at the Family level.
df_family <- tax_glom(df_pruned, taxrank = "Family",NArm =TRUE)

# Extract the OTU table from the family-level object.
otu_family <- otu_table(df_family)

# Ensure that the OTU table orientation is correct. (samples as rows)
if(taxa_are_rows(otu_family)) {
  family_counts <- as.data.frame(t(otu_family))
} else {
  family_counts <- as.data.frame(otu_family)
}

# Update the column names using the Family names from the taxonomy table.
tax_family <- tax_table(df_family)[, "Family"]
colnames(family_counts) <- as.character(tax_family)

# Convert to character vector
family_names <- as.character(tax_family)

# Identify duplicated "uncultured" entries and make them unique
uncultured_indices <- which(family_names == "uncultured")
if (length(uncultured_indices) > 0) {
  family_names[uncultured_indices] <- paste0("uncultured_", seq_along(uncultured_indices))
}

# Assign updated names to the count table
colnames(family_counts) <- family_names

# Inspect the first few rows of the family-level count table.
head(family_counts)
```

## Run Maaslin3
```{r}
#Reload metadata
amylase_metadata <- read.csv("Amylase_metadata_list_ambiguityremoved.csv", row.names = 1)

# Set a seed for reproducibility
set.seed(1)

# Run MaalLin3 (Genus Level)
fit_out <- maaslin3(
  input_data = genus_counts,
  input_metadata = amylase_metadata,
  output = 'genus_maaslin3_output',   # Specify your desired output directory name
  formula = "~ Amylase",  # Replace with actual metadata columns; e.g., "~ diagnosis + age"
  normalization = 'TSS',           # Total Sum Scaling normalization
  transform = 'LOG',               # Log transformation (LOG means log2-transformed relative abundances)
  augment = TRUE,                  # Data augmentation to deal with logistic regression issues
  standardize = TRUE,              # Z-score standardizes continuous metadata variables
  max_significance = 0.05,          # Maximum allowed FDR for significance
  median_comparison_abundance = TRUE,  
  median_comparison_prevalence = FALSE,
  max_pngs = 250,                  # Maximum number of PNG plots to save (if applicable)
  cores = 1                        # Number of CPU cores to use
)

# Run Maaslin3 (Family Level)
fit_out <- maaslin3(
  input_data = family_counts,
  input_metadata = amylase_metadata,
  output = 'family_maaslin3_output',   # Specify your desired output directory name
  formula = "~ Amylase",  
  normalization = 'TSS',           # Total Sum Scaling normalization
  transform = 'LOG',               # Log transformation (LOG means log2-transformed relative abundances)
  augment = TRUE,                  # Data augmentation to deal with logistic regression issues
  standardize = TRUE,              # Z-score standardizes continuous metadata variables
  max_significance = 0.05,          # Maximum allowed FDR for significance
  median_comparison_abundance = TRUE,  
  median_comparison_prevalence = FALSE,
  max_pngs = 250,                  # Maximum number of PNG plots to save (if applicable)
  cores = 1                        # Number of CPU cores to use
)

# -----------------------------------------
# Step 7: Save Outputs as .csv(s)
# -----------------------------------------
# Read the .tsv files
genus_all_results<- read_tsv("C:/path/to/your/directory/genus_maaslin3_output_/all_results.tsv")

family_all_results<- read_tsv("C:/path/to/your/directory/family_maaslin3_output/all_results.tsv")

# Check the first few rows of the data to ensure it was read correctly
print(head(genus_all_results))
print(head(family_all_results))

# Write the tables to .csv files for easy future viewing (optional)
write_csv(genus_all_results,("genus_all_results.csv"))
write_csv(family_all_results,("family_all_results.csv"))

```

## This Script Calculates, Compares, and Plots Alpha Diversity (Observed and Shannon)

```{r}
diversity_metrics <- estimate_richness(df_pruned, measures = c("Shannon", "Observed"))
diversity_metrics$Group <- amylase_metadata$Amylase  # Attaches Amylase Grouping


#Set y axis upper limits and p value label position

obs_upper       <- 500      # observed y-axis maximum
obs_label_y     <- 480      # p-value label position for Observed plot
shannon_upper   <- 6      # Shannon y-axis maximum
shannon_label_y <- 5.79      # p-value label position for Shannon plot

#Sets color palette (colorblind friendly)
obs_fill_values <- c("Low-Intermediate" = "#0072B2", "High" = "#D55E00")
obs_color_values <- obs_fill_values  # same values for fill and color

#Make Observed Diversity Plot

p_observed <- ggplot(diversity_metrics, aes(x = Group, y = Observed, fill = Group)) +
  geom_violin(trim = FALSE, alpha = 0.5, color = "black", width = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.9) +
  scale_fill_manual(name = NULL, values = obs_fill_values) +  # No legend title
  scale_color_manual(name = NULL, values = obs_color_values) +
  labs(title = "Observed Diversity", y = "Num. Observed Taxa") +
  scale_y_continuous(limits = c(0, obs_upper)) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    axis.text.x   = element_blank(),      # Remove x–tick labels
    axis.ticks.x  = element_blank(),      # Remove x–ticks
    axis.title.x  = element_blank(),
    legend.position = "right"
  ) +
  stat_compare_means(
    comparisons = list(c("Low-Intermediate", "High")),
    method      = "wilcox.test",
    label       = "p.signif",
    size        = 6,
    label.y   = 400
  )

#Make Shannon Diversity Plot
p_shannon <- ggplot(diversity_metrics, aes(x = Group, y = Shannon, fill = Group)) +
  geom_violin(trim = FALSE, alpha = 0.5, color = "black", width = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.9) +
  scale_fill_manual(name = NULL, values = obs_fill_values) +
  scale_color_manual(name = NULL, values = obs_color_values) +
  labs(title = "Shannon Diversity", y = "Shannon Diversity Index") +
  scale_y_continuous(limits = c(0, shannon_upper)) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title    = element_text(face = "bold", hjust = 0.5),
    axis.text.x   = element_blank(),      # Remove x–tick labels
    axis.ticks.x  = element_blank(),      # Remove x–ticks
    axis.title.x  = element_blank(),
    legend.position = "right"
  ) +
  stat_compare_means(
    comparisons = list(c("Low-Intermediate", "High")),
    method      = "wilcox.test",
    label       = "p.signif",
    size        = 6,
    label.y   = 5
  )

#Create Combined Plot with Shared Legend
#Remove individual legends
p_observed_no_leg <- p_observed + theme(legend.position = "none")
p_shannon_no_leg  <- p_shannon  + theme(legend.position = "none")

#Arrange the two main plots side by side (ensuring equal widths)
main_plots <- ggarrange(p_observed_no_leg, p_shannon_no_leg,
                        ncol = 2, nrow = 1)

#Create a common x–axis label grob (centered over the two panels)
xlabel_grob <- text_grob("BAL Amylase Group", face = "bold", size = 16, hjust = 0.5)

#Stack the main plots and the common x-axis label vertically
main_with_xlabel <- ggarrange(main_plots, xlabel_grob, ncol = 1, heights = c(10, 1))

# Extract common legend, positioned at the bottom.
common_legend <- get_legend(
  p_shannon + theme(legend.position = "bottom",
                    legend.box = "horizontal")
)

final_plot <- ggarrange(main_with_xlabel, common_legend,
                        ncol = 1, heights = c(10, 1))
```

## Make Relative Abundance Barplot

```{r}

sample_data(df_pruned)$Group <- amylase_metadata$Amylase
ps <- filter_taxa(df_pruned, function(x) sum(x > 3) > (0.2*length(x)), TRUE)
df_rel_abund <- transform_sample_counts(ps, function(x) x / sum(x))
barplot <- phyloseq::plot_bar(df_rel_abund, fill = "Order") + geom_bar(aes(fill = Order), stat = "identity", position = "stack")  + scale_fill_viridis_d(option="mako") + labs(x = "", y = "Relative Abundance\n") + facet_wrap(~ Group, scales = "free") + theme(panel.background = element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(), legend.position="bottom", strip.text = element_text(size=16))
```

## Convert Counts to Relative Abundances and Compute Distances

```{r}

#Convert counts to relative abundance
sample_data(df_pruned)$Group <- amylase_metadata$Amylase
df_pruned_rel <- transform_sample_counts(df_pruned, function(x) x / sum(x))

#Prepare data and compute distances
bray_dist <- phyloseq::distance(df_pruned_rel, method = "bray")
bray_pcoa <- ordinate(df_pruned_rel, method = "PCoA", distance = "bray")

#Extract sample data as a data frame for PERMANOVA
sample_df <- as(sample_data(df_pruned_rel), "data.frame")

#Run PERMANOVA using adonis2
adonis_result <- adonis2(bray_dist ~ Group, data = sample_df)

#Extract and format output statistics
F_val <- round(adonis_result$F[1], 2)
R2_val <- round(adonis_result$R2[1], 2)
p_val  <- adonis_result$`Pr(>F)`[1]
perm_label <- paste0("10% Prevalence Threshold - PERMANOVA: F = ", F_val,
                     ", R² = ", R2_val,
                     ", p = ", format(p_val, digits = 2))

#Calculate percentage variance
pc1 <- round(bray_pcoa$values$Relative_eig[1] * 100, 1)
pc2 <- round(bray_pcoa$values$Relative_eig[2] * 100, 1)


#Produces PCoA with solid line elipses
p_pcoa_solid <- plot_ordination(df_pruned_rel, bray_pcoa, color = "Group") +
  geom_point(size = 3, alpha = 0.9, shape = 16) +
  # Draw 95% confidence ellipses for each group
  stat_ellipse(aes(color = Group), type = "norm",
               linetype = "solid", level = 0.95, linewidth = 1) +
  scale_color_manual(name = "BAL Amylase Group", 
                     values = c("Low-Intermediate" = "#0072B2",
                                "High" = "#D55E00")) +
  labs(title = "Bray–Curtis Dissimilarity (PCoA)",
       x = paste0("PC1 (", pc1, "%)"),
       y = paste0("PC2 (", pc2, "%)")) +
  theme_minimal(base_size = 16) +
  theme(plot.title   = element_text(face = "bold", hjust = 0.5),
        legend.position = "right") +
  # Annotate the plot with PERMANOVA results
  annotate("text", x = Inf, y = Inf, label = perm_label,
           hjust = 1.1, vjust = 1.1, size = 6, fontface = "italic")

#Part B: PC1/PC2 Plots
pc_df <- as.data.frame(bray_pcoa$vectors)
colnames(pc_df)[1:2] <- c("PC1", "PC2")
pc_df$Group <- sample_data(df_pruned_rel)$Group

max_pc1 <- max(pc_df$PC1, na.rm = TRUE)
min_pc1 <- min(pc_df$PC1, na.rm = TRUE)
range_pc1 <- max_pc1 - min_pc1
label_y_pc1 <- 0.45

#For PC2:
max_pc2 <- max(pc_df$PC2, na.rm = TRUE)
min_pc2 <- min(pc_df$PC2, na.rm = TRUE)
range_pc2 <- max_pc2 - min_pc2
label_y_pc2 <- 0.45

#Define color palette
palette <- c("Low-Intermediate" = "#0072B2",
             "High" = "#D55E00")

#PC1 Boxplot (removing individual x axis ticks and labels)
p_pc1 <- ggplot(pc_df, aes(x = Group, y = PC1, fill = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.15, size = 3, alpha = 0.8) +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(title = paste0("PC1 (", pc1, "% variance)"),
       y = "PC1 Score") +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        # Remove individual x-axis ticks, text, and title
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  stat_compare_means(comparisons = list(c("Low-Intermediate", "High")),
                     method = "wilcox.test",
                     label = "p.signif",
                     label.y = label_y_pc1)

#PC2 Boxplot (removing individual x axis ticks and labels)
p_pc2 <- ggplot(pc_df, aes(x = Group, y = PC2, fill = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.15, size = 3, alpha = 0.8) +
  scale_fill_manual(values = palette) +
  scale_color_manual(values = palette) +
  labs(title = paste0("PC2 (", pc2, "% variance)"),
       y = "PC2 Score") +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        # Remove individual x-axis ticks, text, and title
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  stat_compare_means(comparisons = list(c("Low-Intermediate", "High")),
                     method = "wilcox.test",
                     label = "p.signif",
                     label.y = label_y_pc2)


#Arrange the two plots side by side without individual x-axis info
pc_plot <- ggarrange(p_pc1, p_pc2, ncol = 2, nrow = 1)

# Display the final plot
print(pc_plot)
```

## Combine Barplot, Alpha Diversity, PCoA and PC1, PC2 into One Figure

```{r}
pc_plot <- ggarrange(p_pc1, p_pc2, ncol = 2, nrow = 1)
combined_plots <- ggarrange(barplot, final_plot, p_pcoa_solid, pc_plot, labels = c("A", "B", "C", "D"),  ncol = 2, nrow = 2)
print (combined_plots)
```

