---
title: "Amylase Project"
output: html_document
date: "2025-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE)
```

## Generating dataset that will be used for demographics and outcomes (Table 1)

The code chunk below lists the excel sheets that contain clinical and microbiological results from the cohort. After running this code chunk you will have the 'df2' dataframe in which each row corresponds to the first BAL and associated metadata for the patient.

```{r, echo=TRUE}
library(FSA)
library(dplyr)
library(stats)
library(RColorBrewer)
library(ggplot2)
library(viridis)
library(hrbrthemes)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(tibble)
library(modelsummary)
library(reshape)
library(tidyr)
library("survival")
library("survminer")
library(patchwork)
library(fabricatr)
library(mice)
library(visdat)
library(fabricatr)
library(readxl)

#Select BAL samples with an amylase value
CDdataset <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/CDdataset.xlsx")
EDW_bal_results <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/EDW_bal_results.xlsx")
EDW_basic_endpoints <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/EDW_basic_endpoints.xlsx")
Antibiotics <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/Antibiotics.xlsx")
SCRIPT_Cardiac_Arrest <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/SCRIPT_Cardiac_Arrest.xlsx")
Amylase <- subset(CDdataset, !is.na(amylase))

#Select the first BAL from each patient
Amylase <- Amylase[grep("BAL-00", Amylase$bal_barcode), ]

#Add a column for Mortality
Amylase <- Amylase %>% mutate(Mortality = case_when(discharge_disposition_name == 'Died' ~ "1"))
Amylase <- Amylase %>% mutate(Mortality = ifelse(is.na(Mortality), 0, Mortality))

#Merge 'Amylase' dataframe and 'EDW_bal_results' since EDW_bal_results has the specific culture results
Merged_df = merge(Amylase, EDW_bal_results, by.x=c("patient_ir_id", "bal_dt"), by.y=c("patient_ir_id", "bal_collection_date"))

#Remove duplicate rows based on patient IRID
#Remove BALs that fall outside the first intubation window
df <- Merged_df[!duplicated(Merged_df$patient_ir_id),]

#Historgram of Amylase distribution (Supplemental Figure 2)
p <- ggplot(df, aes(log(amylase))) + geom_histogram(aes(y = ..density..), color = "#000000", fill = "#0099F8") + geom_density(color = "#000000", fill = "lightblue", alpha = 0.6) + theme_bw()
p + labs(title="Density Plot of Distribution of BAL Amylase Levels", y= "Density", x="Log(BAL Amylase)") + theme(plot.title = element_text(hjust = 0.5))

#Add column for Quartile and split quartiles
df <- df %>% mutate(Quartile = split_quantile(amylase, 4))

#Add column called 'group' to compare Quartiles 1-3 to Quartile 4
df$Quartile <- as.numeric(df$Quartile)
df <- df %>% mutate(group = case_when(Quartile <= 3 ~ "Low-Intermediate BAL Amylase", Quartile > 3 ~ "High BAL Amylase"))

#Subset df for Table 1 Data
df2 <- df %>% select(patient_ir_id, day_of_hospitalization, day_of_intubation, bal_dt, pt_study_id, norepinephrine_avg, edw_adm_age, bmi, gender, ethnicity, pt_immunocomp, smoking_history, myocardial_infarction, congestive_heart_failure, cerebrovascular_disease, dementia, peptic_ulcer_disease, liver_disease, diabetes, hemiplagia_or_paraplegia, charlson_comorbidity_index, hospital_los_days, total_icu_los_days, cumulative_intubation_days, discharge_datetime, discharge_disposition_name, Mortality, temperature_max, norepinephrine_avg, fio2_avg, po2_fio2_ratio_avg, wbc_max, amylase, bal_wbc_count, bal_pct_neutro, gram_stain_report_organisms, organism_1_quantity, organism_1_name, organism_2_quantity, organism_2_name, organism_3_quantity, organism_3_name, organism_4_quantity, organism_4_name, organism_5_quantity, organism_5_name, organism_6_quantity, organism_6_name, episode_category, episode_etiology, Quartile, group)

```


## Visualize missing data (Supplement Figure 2)

```{r pressure, echo=TRUE}
#Visualize missing data (Supplemental Figure 1)
vis_miss(df2)
```

## Impute missing data 
First replace 'NA' in the BAL culture columns with the text 'NA' since there is nothing to impute in these cells, NA means no growth/normal flora
```{r, echo=TRUE}
df2[35:49][is.na(df2[35:49])] <- 'NA'
df2_imp <- mice(df2,m=5,maxit=50,meth="cart",seed=500)
df2_imp <- complete(df2_imp, 1)

#Add Cardiac_arrest data from RedCap
df2_imp <- merge(df2_imp, SCRIPT_Cardiac_Arrest, by.x = "pt_study_id", by.y = "pt_study_id")

#Visualize missing data again to ensure there are no longer any missing datapoints
vis_miss(df2_imp)
```

#Rename organisms to genus level

```{r, echo=FALSE}
df2_imp$organism_1_name[df2_imp$organism_1_name=="Acinetobacter baumannii complex"] <- "Acinetobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Neisseria meningitidis"] <- "Neisseria"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans streptococcus #3"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella (Enterobacter) aerogenes"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Pseudomonas aeruginosa (Encapsulated Strain)"] <- "Pseudomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Burkholderia cepacia complex"] <- "Burkholderia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Burkholderia cepacia complex (presumptive)"] <- "Burkholderia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes (presumptive)"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Elizabethkingia meningoseptica"] <- "Elizabethkingia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter aerogenes"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter cloacae complex"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter cloacae complex (ESBL Positive)"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli (ESBL Positive) "] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli"] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Serratia liquefaciens"] <- "Serratia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus mitis oralis group"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterococcus faecium (Vancomycin Resistant)"] <- "Enterococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli (ESBL Positive)"] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Hafnia alvei"] <- "Hafnia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella pneumoniae (ESBL Positive)"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Morganella morganii"] <- "Morganella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Methicillin-Resistant Staphylococcus aureus"] <- "Staphylococcus aureus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Stenotrophomonas maltophilia"] <- "Stenotrophomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Stomatococcus species"] <- "Stomatococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus agalactiae (Group B)"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus pseudopneumoniae"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida albicans"] <- "Candida"


df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group C"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group F"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group G"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans streptococcus #2"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecium"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecium (Vancomycin Resistant)"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Escherichia coli (ESBL Positive)"] <- "Escherichia"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Klebsiella pneumoniae (ESBL Positive)"] <- "Klebsiella"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Proteus mirabilis"] <- "Proteus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Methicillin-Resistant Staphylococcus aureus"] <- "Staphylococcus aureus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Stomatococcus species"] <- "Stomatococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Streptococcus pneumoniae"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Streptococcus pseudopneumoniae"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Yeast, Not Cryptococcus Species #2"] <- "Yeast"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida albicans"] <- "Candida"

df2_imp$organism_3_name[df2_imp$organism_3_name=="Staphylococcus aureus #2"] <- "Staphylococcus aureus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Lactobacillus species #2"] <- "Lactobacillus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans streptococcus #2"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group C"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group F"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group G"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Pseudomonas aeruginosa #2 (Encapsulated Strain)"] <- "Pseudomonas"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Achromobacter species"] <- "Achromobacter "
df2_imp$organism_3_name[df2_imp$organism_3_name=="Acinetobacter baumannii complex Cefiderocol MIC = 0.5, No Interpertation"] <- "Acinetobacter"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Enterococcus faecium"] <- "Enterococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Granulicatella adiacens"] <- "Granulicatella"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Proteus mirabilis"] <- "Proteus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="streptococcus"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida albicans"] <- "Candida"

df2_imp$organism_4_name[df2_imp$organism_4_name=="Arcanobacterium haemolyticum"] <- "Arcanobacterium"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Enterococcus raffinosus"] <- "Enterococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Hafnia alvei"] <- "Hafnia"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida albicans"] <- "Candida"

df2_imp$organism_5_name[df2_imp$organism_5_name=="Acinetobacter ursingii"] <- "Acinetobacter"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Haemophilus Streptococcus constellatus "] <- "Haemophilus"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Streptococcus constellatus"] <- "Streptococcus"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida albicans"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida albicans"] <- "Candida"
```

##Get summary statistics for Low-Intermediate vs High BAL fluid amylase groups (Table 1)

```{r, echo = TRUE}
#Find culture-negative samples which are the ones where the organism 1 name is NA, meaning no bacteria grew on culture
df2_imp <- df2_imp %>% mutate(Culture_negative = case_when(organism_1_name == 'NA' ~ "Yes"))
df2_imp$Culture_negative <- df2_imp$Culture_negative %>% replace_na('No')

#Find polymicrobial samples, meaning the organism_1_name and organism_2_name columns are not blank
df2_imp <- df2_imp %>% mutate(Polymicrobial = case_when(organism_1_name != 'NA' & organism_2_name != 'NA' ~ "Yes"))
df2_imp$Polymicrobial <- df2_imp$Polymicrobial %>% replace_na('No')

#Generate Table 1
df2_imp %>% select(group, amylase, edw_adm_age, bmi, gender, charlson_comorbidity_index, Cardiac_arrest, pt_immunocomp, peptic_ulcer_disease, cerebrovascular_disease, hemiplagia_or_paraplegia, dementia, episode_category, Culture_negative, Polymicrobial, wbc_max, bal_wbc_count, bal_pct_neutro, hospital_los_days, total_icu_los_days, Mortality) %>% tbl_summary(by= group, statistic = list(all_categorical() ~ "{n} / {N} ({p}%)")) %>% add_p() %>% add_q()

#Find cases where Gram stain is positive for culture negatives
#The output is not significant and not corrected for multiple comparisons
cx_neg <- df2_imp %>% filter(is.na(organism_1_name))
cx_neg %>% count(gram_stain_report_organisms, group, sort = TRUE)
gs_pos <- data.frame("gs_pos" = c(19,7), "No_gs_pos" = c(233,39), row.names = c("Low-Intermediate", "High"))
fisher.test(gs_pos)

```


