---
title: "Amylase Project"
output: html_document
date: "2025-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, echo = TRUE)
```

## Generating dataset that will be used for demographics and outcomes (Table 1)

The code chunk below lists the excel sheets that contain clinical and microbiological results from the cohort. After running this code chunk you will have the 'df2' dataframe in which each row corresponds to the first BAL and associated metadata for the patient.

```{r, echo=TRUE}
library(FSA)
library(dplyr)
library(stats)
library(RColorBrewer)
library(ggplot2)
library(viridis)
library(hrbrthemes)
library(tidyverse)
library(ggpubr)
library(lubridate)
library(tibble)
library(modelsummary)
library(reshape)
library(tidyr)
library("survival")
library("survminer")
library(patchwork)
library(fabricatr)
library(mice)
library(visdat)
library(fabricatr)
library(readxl)
library(gtsummary)

#Select BAL samples with an amylase value
CDdataset <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/CDdataset.xlsx")
EDW_bal_results <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/EDW_bal_results.xlsx")
EDW_basic_endpoints <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/EDW_basic_endpoints.xlsx")
Antibiotics <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/Antibiotics.xlsx")
SCRIPT_Cardiac_Arrest <- read_excel("C:/Users/cononye/Desktop/RStudio/Amylase/SCRIPT_Cardiac_Arrest.xlsx")

#Rename dataset and select all BALs that have an amylase value measured
Amylase <- subset(CDdataset, !is.na(amylase))

#Select the first BAL from each patient
#Note that this may not necessarily correspond to the first enrollment intubation as not all enrollment intubations had an amylase value measured
Amylase <- Amylase[grep("BAL-00", Amylase$bal_barcode), ]

#Add a column for Mortality
Amylase <- Amylase %>% mutate(Mortality = case_when(discharge_disposition_name == 'Died' ~ "1"))
Amylase <- Amylase %>% mutate(Mortality = ifelse(is.na(Mortality), 0, Mortality))

#Merge 'Amylase' dataframe and 'EDW_bal_results' since EDW_bal_results has the specific culture results
Merged_df = merge(Amylase, EDW_bal_results, by.x=c("patient_ir_id", "bal_dt"), by.y=c("patient_ir_id", "bal_collection_date"))

#Remove duplicate rows based on patient IRID
df <- Merged_df[!duplicated(Merged_df$patient_ir_id),]

#Historgram of Amylase distribution (Supplemental Figure 2)
p <- ggplot(df, aes(log(amylase))) + geom_histogram(aes(y = ..density..), color = "#000000", fill = "#0099F8") + geom_density(color = "#000000", fill = "lightblue", alpha = 0.6) + theme_bw()
p + labs(title="Density Plot of Distribution of BAL Amylase Levels", y= "Density", x="Log(BAL Amylase)") + theme(plot.title = element_text(hjust = 0.5))

#Add column for Quartile and split quartiles
#Get number of samples in each quartile
df <- df %>% mutate(Quartile = split_quantile(amylase, 4))
table(df$Quartile)

#Get median and IQR for amylase
tapply(df$amylase, df$Quartile, summary)

#Add column called 'group' to compare Quartiles 1-3 to Quartile 4
df$Quartile <- as.numeric(df$Quartile)
df <- df %>% mutate(group = case_when(Quartile <= 3 ~ "Low-Intermediate BAL Amylase", Quartile > 3 ~ "High BAL Amylase"))


#Subset df for Table 1 Data
df2 <- df %>% select(patient_ir_id, day_of_hospitalization, day_of_intubation, bal_dt, pt_study_id, norepinephrine_avg, edw_adm_age, bmi, gender, ethnicity, pt_immunocomp, smoking_history, myocardial_infarction, congestive_heart_failure, cerebrovascular_disease, dementia, peptic_ulcer_disease, liver_disease, diabetes, hemiplagia_or_paraplegia, charlson_comorbidity_index, hospital_los_days, total_icu_los_days, cumulative_intubation_days, discharge_datetime, discharge_disposition_name, Mortality, temperature_max, norepinephrine_avg, fio2_avg, po2_fio2_ratio_avg, wbc_max, amylase, bal_wbc_count, bal_pct_neutro, gram_stain_report_organisms, organism_1_quantity, organism_1_name, organism_2_quantity, organism_2_name, organism_3_quantity, organism_3_name, organism_4_quantity, organism_4_name, organism_5_quantity, organism_5_name, organism_6_quantity, organism_6_name, episode_category, episode_etiology, Quartile, group)

df2$po2_fio2_ratio_avg <- as.numeric(df2$po2_fio2_ratio_avg)
```


## Visualize missing data (Supplement Figure 2)
```{r pressure, echo=FALSE}
#First replace 'NA' in the BAL culture columns with the text 'NA' since there is nothing to impute in these cells, NA means no growth/normal flora.
df2[35:49][is.na(df2[35:49])] <- 'NA'

#Visualize missing data (Supplemental Figure 2)
vis_miss(df2)

```

## Impute missing data 
```{r, echo=FALSE}
df2_imp <- mice(df2,m=5,maxit=50,meth="cart",seed=500)
df2_imp <- complete(df2_imp, 1)

#Add Cardiac_arrest data from RedCap
df2_imp <- merge(df2_imp, SCRIPT_Cardiac_Arrest, by.x = "pt_study_id", by.y = "pt_study_id")

#Visualize missing data again to ensure there are no longer any missing datapoints
vis_miss(df2_imp)
```

#Rename organisms to genus level

```{r, echo=FALSE}
df2_imp$organism_1_name[df2_imp$organism_1_name=="Acinetobacter baumannii complex"] <- "Acinetobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Neisseria meningitidis"] <- "Neisseria"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans streptococcus #3"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella (Enterobacter) aerogenes"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Pseudomonas aeruginosa (Encapsulated Strain)"] <- "Pseudomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Burkholderia cepacia complex"] <- "Burkholderia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Burkholderia cepacia complex (presumptive)"] <- "Burkholderia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes (presumptive)"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Elizabethkingia meningoseptica"] <- "Elizabethkingia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter aerogenes"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter cloacae complex"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterobacter cloacae complex (ESBL Positive)"] <- "Enterobacter"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli (ESBL Positive) "] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli"] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Chryseobacterium indologenes"] <- "Chryseobacterium"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Serratia liquefaciens"] <- "Serratia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus mitis oralis group"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Enterococcus faecium (Vancomycin Resistant)"] <- "Enterococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Escherichia coli (ESBL Positive)"] <- "Escherichia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Hafnia alvei"] <- "Hafnia"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella pneumoniae (ESBL Positive)"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Morganella morganii"] <- "Morganella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Methicillin-Resistant Staphylococcus aureus"] <- "Staphylococcus aureus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Stenotrophomonas maltophilia"] <- "Stenotrophomonas"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Stomatococcus species"] <- "Stomatococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus agalactiae (Group B)"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Streptococcus pseudopneumoniae"] <- "Streptococcus"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_1_name[df2_imp$organism_1_name=="Candida albicans"] <- "Candida"


df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group C"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group F"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, Group G"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Beta Hemolytic Streptococci, not Group A, B, C, D, F, or G"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans streptococcus #2"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecium"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Enterococcus faecium (Vancomycin Resistant)"] <- "Enterococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Escherichia coli (ESBL Positive)"] <- "Escherichia"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Klebsiella pneumoniae (ESBL Positive)"] <- "Klebsiella"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Proteus mirabilis"] <- "Proteus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Methicillin-Resistant Staphylococcus aureus"] <- "Staphylococcus aureus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Stomatococcus species"] <- "Stomatococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Streptococcus pneumoniae"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Streptococcus pseudopneumoniae"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Yeast, Not Cryptococcus Species #2"] <- "Yeast"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_2_name[df2_imp$organism_2_name=="Candida albicans"] <- "Candida"

df2_imp$organism_3_name[df2_imp$organism_3_name=="Staphylococcus aureus #2"] <- "Staphylococcus aureus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Lactobacillus species #2"] <- "Lactobacillus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans streptococcus #2"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans Streptococcus Group"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group C"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group F"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Beta Hemolytic Streptococci, Group G"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Pseudomonas aeruginosa #2 (Encapsulated Strain)"] <- "Pseudomonas"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Achromobacter species"] <- "Achromobacter "
df2_imp$organism_3_name[df2_imp$organism_3_name=="Acinetobacter baumannii complex Cefiderocol MIC = 0.5, No Interpertation"] <- "Acinetobacter"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Citrobacter koseri"] <- "Citrobacter"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Enterococcus faecium"] <- "Enterococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Granulicatella adiacens"] <- "Granulicatella"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Positive"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Pseudomonas aeruginosa"] <- "Pseudomonas"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Proteus mirabilis"] <- "Proteus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Lactobacillus species"] <- "Lactobacillus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="streptococcus"] <- "Streptococcus"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_3_name[df2_imp$organism_3_name=="Candida albicans"] <- "Candida"

df2_imp$organism_4_name[df2_imp$organism_4_name=="Arcanobacterium haemolyticum"] <- "Arcanobacterium"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Enterococcus faecalis"] <- "Enterococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Enterococcus raffinosus"] <- "Enterococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Hafnia alvei"] <- "Hafnia"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Klebsiella pneumoniae"] <- "Klebsiella"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Neisseria species"] <- "Neisseria"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Serratia marcescens"] <- "Serratia"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Viridans streptococcus"] <- "Streptococcus"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Yeast, Not Cryptococcus Species"] <- "Yeast"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_4_name[df2_imp$organism_4_name=="Candida albicans"] <- "Candida"

df2_imp$organism_5_name[df2_imp$organism_5_name=="Acinetobacter ursingii"] <- "Acinetobacter"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Haemophilus Streptococcus constellatus "] <- "Haemophilus"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Corynebacterium species"] <- "Corynebacterium"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Streptococcus constellatus"] <- "Streptococcus"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Haemophilus influenzae Beta Lactamase Negative"] <- "Haemophilus"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_5_name[df2_imp$organism_5_name=="Candida albicans"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida glabrata"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida tropicalis"] <- "Candida"
df2_imp$organism_6_name[df2_imp$organism_6_name=="Candida albicans"] <- "Candida"
```

##Get additional descriptive variables like number of culture negative samples, number of polymicrobial samples and re-name columns.
```{r}
#Find culture-negative samples which are the ones where the organism 1 name is NA, meaning no bacteria grew on culture
df2_imp <- df2_imp %>% mutate(Culture_negative = case_when(organism_1_name == 'NA' ~ "Yes"))
df2_imp$Culture_negative <- df2_imp$Culture_negative %>% replace_na('No')

#Find polymicrobial samples, meaning the organism_1_name and organism_2_name columns are not blank
df2_imp <- df2_imp %>% mutate(Polymicrobial = case_when(organism_1_name != 'NA' & organism_2_name != 'NA' ~ "Yes"))
df2_imp$Polymicrobial <- df2_imp$Polymicrobial %>% replace_na('No')

#Rename columns for Table 1
df2_imp$Mortality[df2_imp$Mortality=="0"] <- "Alive"
df2_imp$Mortality[df2_imp$Mortality=="1"] <- "Deceased"
df2_imp$hemiplagia_or_paraplegia[df2_imp$hemiplagia_or_paraplegia=="0"] <- "No"
df2_imp$hemiplagia_or_paraplegia[df2_imp$hemiplagia_or_paraplegia=="2"] <- "Yes"
df2_imp$episode_category[df2_imp$episode_category=="CAP"] <- "Community Acquired Pneumonia"
df2_imp$episode_category[df2_imp$episode_category=="HAP"] <- "Hospital Acquired Pneumonia" 
df2_imp$episode_category[df2_imp$episode_category=="NPC"] <- "Non Pneumonia Control"
df2_imp$episode_category[df2_imp$episode_category=="VAP"] <- "Ventilator Associated Pneumonia"

list_of_labels <- list(amylase="BAL Fluid Amylase Level, U/L", edw_adm_age ="Age, years", bmi = "Body Mass Index, kg/m2", gender="Gender, n (%)", charlson_comorbidity_index= "Charlson Comorbidity Index", Cardiac_arrest="Cardiac Arrest Prior to Intubation, n (%)", pt_immunocomp="Immunocompromised, n (%)", peptic_ulcer_disease="Peptic Ulcer Disease, n (%)", cerebrovascular_disease="Cerebrovascular Disease, n (%)", hemiplagia_or_paraplegia="Hemiplegia or Paraplegia, n (%)", Culture_negative="Negative BAL Culture, n (%)", dementia="Dementia, n (%)", episode_category="Pneumonia Type, n (%)", Polymicrobial="Polymicrobial BAL Culture, n (%)", wbc_max="Peripheral Blood Leukocyte Count, cells/microliter", bal_wbc_count="BAL Leukocyte Count, cells/microliter", bal_pct_neutro="BAL Neutrophil Percentage", hospital_los_days="Hospital Length of Stay, days", total_icu_los_days="Total ICU Length of Stay, days", Mortality="Hospital Discharge Status, n (%)")
```

##Get summary statistics for Low-Intermediate vs High BAL fluid amylase groups (Table 1)
```{r, echo = TRUE}
tapply(df2_imp$amylase, df2_imp$group, summary)
Table1 <- df2_imp
Table1 %>% select(group, amylase, edw_adm_age, bmi, gender, charlson_comorbidity_index, Cardiac_arrest, pt_immunocomp, peptic_ulcer_disease, cerebrovascular_disease, hemiplagia_or_paraplegia, dementia, episode_category, Culture_negative, Polymicrobial, wbc_max, bal_wbc_count, bal_pct_neutro, hospital_los_days, total_icu_los_days, Mortality) %>% tbl_summary(by= group, statistic = list(all_continuous() ~ "{median} ({p25}, {p75})", all_categorical() ~ "{n} ({p}%)"), label=list_of_labels, ) %>% add_p(pvalue_fun = label_style_pvalue(digits = 2))
```

```{r, echo=FALSE}
#Find cases where Gram stain is positive for culture negatives
#The output is significant
cx_neg <- df2_imp %>% mutate(Gram_stain = case_when(gram_stain_report_organisms == 
'Budding Yeast Cells with Pseudohyphae Seen' ~ "Negative", 
gram_stain_report_organisms == 'Budding yeast seen' ~ "Negative", 
gram_stain_report_organisms == 'Budding yeast with pseudohyphae seen' ~ "Negative", gram_stain_report_organisms == 'NA' ~ "Negative", 
gram_stain_report_organisms == 'No organisms seen' ~ "Negative", 
gram_stain_report_organisms == 'No Organisms Seen.' ~ "Negative", 
gram_stain_report_organisms == 'Yeast' ~ "Negative", 
gram_stain_report_organisms == 'Yeast Seen' ~ "Negative",))

cx_neg$Gram_stain = cx_neg$Gram_stain %>% replace_na("Positive")
cx_neg %>% select(group, Gram_stain) %>% tbl_summary(by= group, statistic = list(all_categorical() ~ "{n} / {N} ({p}%)")) %>% add_p(pvalue_fun = function(x) style_pvalue(x, digits = 2))
```

##Evaluate significant demographics using BAL amylase as a continuous variable 
```{r, echo=FALSE}
df2_imp$Quartile <- as.factor(df2_imp$Quartile)
a <- ggplot(df2_imp, aes(x=log(amylase), y= log(bal_wbc_count))) + scale_color_viridis_d()
a <- a + geom_point(aes(color=Quartile), size=4) + geom_smooth(method = "loess") + stat_cor(method = "pearson", label.x = 3, label.y = 12, size=4) + xlab("Log(BAL Amylase)") + ylab("BAL WBC Count") + theme_bw() + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), legend.position='none')

b <- ggplot(df2_imp, aes(x=log(amylase), y=edw_adm_age)) + scale_color_viridis_d()
b <- b + geom_point(aes(color=Quartile), size=4) + geom_smooth(method = "loess") + stat_cor(method = "pearson", label.x = 3, label.y = 105, size=4) + xlab("Log(BAL Amylase)") + ylab("Age") + theme_bw() + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y=element_text(size=12))

c <- ggplot(df2_imp, aes(x=log(amylase), y= bal_pct_neutro)) + scale_color_viridis_d()
c <- c + geom_point(aes(color=Quartile), size=4) + geom_smooth(method = "loess") + stat_cor(method = "pearson", label.x = 3, label.y = 110, size=4) + xlab("Log(BAL Amylase)") + ylab("BAL % Neutrophils") + theme_bw() + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), legend.position='none')

d <- ggplot(df2_imp, aes(x=log(amylase), y= bmi)) + scale_color_viridis_d()
d <- d + geom_point(aes(color=Quartile), size=4) + geom_smooth(method = "loess") + stat_cor(method = "pearson", label.x = 3, label.y = 75, size=4) + xlab("Log(BAL Amylase)") + ylab("Body Mass Index") + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y=element_text(size=12), legend.position='none')

#The statistic here is NS after adjustment for multiple comparisons that was done in the code chunks below
e <- ggplot(df2_imp, aes(x=gender, y=log(amylase))) + geom_boxplot(aes(fill=gender)) + scale_fill_viridis_d()
e <- e + xlab("Gender") + ylab("Log(BAL Amylase)") + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), axis.text.x=element_text(size=12), legend.position='none') + geom_signif(comparisons=list(c("Male", "Female")), map_signif_level = TRUE, y_position = 9.5, tip_length = 0,  textsize = 4, annotations="NS.") + scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

f <- ggplot(df2_imp, aes(x =Cardiac_arrest, y=log(amylase))) + geom_boxplot(aes(fill=Cardiac_arrest)) + scale_fill_viridis_d()
f <- f + xlab("Cardiac Arrest Prior to BAL") + ylab("Log(BAL Amylase)") + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), axis.text.x=element_text(size=12), legend.position='none') + geom_signif(comparisons=list(c("Yes", "No")), map_signif_level = TRUE, y_position = 9.5, tip_length = 0,  textsize = 4) + scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

g <- ggplot(df2_imp, aes(x=Culture_negative, y=log(amylase))) + geom_boxplot(aes(fill=Culture_negative)) + scale_fill_viridis_d()
g <- g + xlab("Culture Negative BAL") + ylab("Log(BAL Amylase)") + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), axis.text.x=element_text(size=12), legend.position='none') + geom_signif(comparisons=list(c("Yes", "No")), map_signif_level = TRUE, y_position = 9.5, tip_length = 0,  textsize = 4) + scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

h <- ggplot(df2_imp, aes(x=Polymicrobial, y=log(amylase))) + geom_boxplot(aes(fill=Polymicrobial)) + scale_fill_viridis_d()
h <- h + xlab("Polymicrobial Culture") + ylab("Log(BAL Amylase)") + theme_bw() + theme(axis.title.x = element_text(size=12), axis.title.y = element_text(size=12), axis.text.x=element_text(size=12), legend.position='none') + geom_signif(comparisons=list(c("Yes", "No")), map_signif_level = TRUE, y_position = 9.5, tip_length = 0,  textsize = 4) + scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

Figure1 <- ggarrange(b, d, a, c, e, f, g, h,  labels = c("A", "B", "C", "D", "E", "F", "G", "H"), ncol = 2, nrow = 4)

print(Figure1)
```

## Get statistics to determine the presence of a linear relationship.
The p value reflects whether or not a linear relationship is present between the two continuous variables for the scatter plots. The p value for the boxplots reflects a difference between median amylase levels based on the presence or absence of a categorical variable (like cardiac arrest, gender).

```{r, echo=FALSE}
#Get statistics for Figure 1
correlation_test_wbc <- cor.test(log(df2_imp$amylase), df2_imp$bal_wbc_count, method = "pearson")
print(correlation_test_wbc)

correlation_test_pmn <- cor.test(log(df2_imp$amylase), df2_imp$bal_pct_neutro, method = "pearson")
print(correlation_test_pmn)
p_value <- correlation_test_pmn$p.value

correlation_test_age <- cor.test(log(df2_imp$amylase), df2_imp$edw_adm_age, method = "pearson")
print(correlation_test_age)
p_value <- correlation_test_age$p.value

correlation_test_bmi <- cor.test(log(df2_imp$amylase), df2_imp$bmi, method = "pearson")
print(correlation_test_bmi)
p_value <- correlation_test_bmi$p.value

gender <- wilcox.test(log(df2_imp$amylase) ~ df2_imp$gender)
print(gender)

cardiac_arrest <- wilcox.test(log(df2_imp$amylase) ~ df2_imp$Cardiac_arrest)
print(cardiac_arrest)

culture_negative <- wilcox.test(log(df2_imp$amylase) ~ df2_imp$Culture_negative)
print(culture_negative)

Polymicrobial <- wilcox.test(log(df2_imp$amylase) ~ df2_imp$Polymicrobial)
print(Polymicrobial)  
```

## Adjust Figure 1 Variables for Multiple Comparisons
```{r, echo = FALSE}
#Create a table with label and p values
p_value_table <- data.frame(Name=c("BAL_wbc_count", "BAL_pct_pmn", "Age", "BMI", "Gender", "Cardiac_arrest", "Culture_negative", "Polymicrobial"), p_values=c(0.00000000000009721, 0.00000000000000022, 0.000002346, 0.002751, 0.03449, 0.11, 0.0000000000001091, 0.0000000007334))
                            
#Enter each p value from the above statistical tests
p_values=c(0.00000000000009721, 0.00000000000000022, 0.000002346, 0.002751, 0.03449, 0.11, 0.0000000000001091, 0.0000000007334)
p_adjusted <- p.adjust(p_values, method = "bonferroni")
p_value_table$p_adjusted <- p_adjusted <- p.adjust(p_values, method = "bonferroni")
print(p_value_table)
```

