---
title: "Time_to_Event_Curves"
output: html_document
date: "2025-05-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Generate time-to-event curves for the time to extubation

```{r, echo=FALSE}
#Add columns for time from intubation to bal, duration of intubation, and time from extubation to discharge
df2_imp_extubation <- EDW_basic_endpoints[EDW_basic_endpoints$pt_study_id %in% c(df2_imp$pt_study_id),]
Extubation_df <- merge(df2_imp, df2_imp_extubation, by=("pt_study_id"))

#BALs from the patient's enrollment intubation 
Extubation_df2 <- Extubation_df
Extubation_df2$intubation_duration <- time_length(Extubation_df2$First_intub_stop - Extubation_df2$First_intub_start, unit="days")
Extubation_df2$intubation_to_bal <- time_length(Extubation_df2$bal_dt - Extubation_df2$First_intub_start, unit="days")
Extubation_df2$extubation_to_discharge <- time_length(Extubation_df2$discharge_datetime.x - Extubation_df2$First_intub_stop, unit="days")
Extubation_df3 <- Extubation_df2[!Extubation_df2$pt_study_id %in% c("1025", "1053", "1103", "1113", "1137", "1155", "1186", "1358", "1427", "1581", "1585", "1651", "1660", "1656", "1676"),]

#Bals that were done during a second or third intubation
Second_intubation <- Extubation_df[Extubation_df$pt_study_id %in% c("1053", "1103", "1113", "1137", "1155", "1186", "1358", "1581", "1585", "1660", "1656", "1676"),]
Third_intubation <- Extubation_df[Extubation_df$pt_study_id %in% c("1025", "1427", "1651"),]

Second_intubation$intubation_duration <- time_length(Second_intubation$Second_intub_stop - Second_intubation$Second_intub_start, unit="days")
Third_intubation$intubation_duration <- time_length(Third_intubation$Third_intub_stop - Third_intubation$Third_intub_start, unit="days")

Second_intubation$intubation_to_bal <- time_length(Second_intubation$bal_dt - Second_intubation$Second_intub_start, unit="days")
Third_intubation$intubation_to_bal <- time_length(Third_intubation$bal_dt - Third_intubation$Third_intub_start, unit="days")

Second_intubation$extubation_to_discharge <- time_length(Second_intubation$discharge_datetime.x - Second_intubation$Second_intub_stop, unit="days")
Third_intubation$extubation_to_discharge <- time_length(Third_intubation$discharge_datetime.x - Third_intubation$Third_intub_stop, unit="days")

Extubation_df3$Intubationnumber <- c("First Intubation")
Second_intubation$Intubationnumber <- c("Second Intubation")
Third_intubation$Intubationnumber <- c("Third Intubation")

#Combine all samples 
All_samples <- rbind(Extubation_df3, Second_intubation, Third_intubation)

#Find duration of intubation for both groups
tapply(All_samples$intubation_duration, All_samples$group, summary)
wilcox.test(intubation_duration ~ group, data=All_samples)

#Generate density plot
density_plot <- ggplot(All_samples, aes(x=intubation_duration, fill=group)) + geom_density(alpha=0.4)
dns_plot <- density_plot + scale_fill_brewer(palette="Dark2") + labs(x="Days of Intubation (For Intubation Associated with the BAL)", y="Density", title = "Density Plot of Intubation Duration") + theme(legend.position="bottom", axis.title.x = element_text(size=14), axis.text.y = element_text(size=14), axis.title.y = element_text(size=14), legend.title = element_text(size=14), legend.text = element_text(size=14) , plot.title=element_text(face="bold", hjust = 0.5), axis.text.x=element_text(size=14)) + guides(fill=guide_legend(title="Group")) + xlim(0,75)

print(dns_plot)
```

##Time to Event Curves

Create a dataframe where the event is time to extubation and the censoring event is patients who died less than 48 hours after extubation. 
```{r, echo=FALSE, warning=FALSE, message=FALSE}

#Create dataframe
Time_to_event <- All_samples %>% select(pt_study_id, amylase, group, intubation_duration, extubation_to_discharge, Mortality, norepinephrine_avg, edw_adm_age, Cardiac_arrest, gender, bmi)

#Create a column called 'status' where 0 is the patient who was censored before experiencing the event, and 1 was the event (extubation)
#We are censoring patients who died and had a time from extubation to discharge < 2 days, meaning they either died on the vent or possibly palliatively extubated
Time_to_event <- Time_to_event %>% mutate(status=case_when(Mortality=="Deceased" & extubation_to_discharge < 2 ~ "0"))

#Status '1' are those who were extubated and alive
Time_to_event$status[is.na(Time_to_event$status)] <- '1'
Time_to_event$status <- as.numeric(Time_to_event$status)

#Change mortality to numeric value
Time_to_event$Mortality[Time_to_event$Mortality=="Alive"] <- 0
Time_to_event$Mortality[Time_to_event$Mortality=="Deceased"] <- 1

#Generate time-to-event curve for time to extubation
fit <- survfit(Surv(intubation_duration, status) ~ group, data=Time_to_event)
print(fit)
km_curve <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE, risk.table = TRUE, risk.table.col = "strata", linetype = "strata", surv.median.line = "hv", ggtheme = theme_bw(), palette = c("Dark2"), xlim=c(0,45), font.main = c(16, "bold"), font.x = c(14, "bold"), font.y = c(14, "bold"), font.tickslab = c(12, "plain"), xlab = "Time to Extubation in Days", ylab= "Probability of Remaining Intubated", font.legend=c(12, "bold"), title="", break.x.by=5, legend.title="BAL Amylase Group", legend.labs=c("High","Low-Intermediate"), legend.position="bottom")

print(km_curve)

#Table showing each individual event
d <- data.frame(time = fit$time, n.risk = fit$n.risk, n.event = fit$n.event, n.censor = fit$n.censor, surv = fit$surv, upper = fit$upper, lower = fit$lower)

```

## Log rank test for p value for KM curve, Cox proportional hazards

```{r, echo=FALSE}
#Get log rank test p value from km curve
surv_pvalue(fit, combine = TRUE, get_coord = TRUE)

##Cox proportional hazards with adjusting for confounders
cox <- coxph(Surv(intubation_duration, status) ~ group + edw_adm_age + bmi + norepinephrine_avg + gender + Cardiac_arrest + Mortality, data = Time_to_event)
summary(cox)
```

## Sensitivity Analysis for All Samples Using Different Amylase Thresholds

```{r, echo=FALSE}
#Sensitivity analysis using different BAL thresholds
Time_to_event2 <- Time_to_event %>% mutate(Amylase_250=case_when(amylase < 250 ~ "Low-Intermediate BAL Amylase", amylase >= 250 ~ "High BAL Amylase"))
Time_to_event2 <- Time_to_event2 %>% mutate(Amylase_400=case_when(amylase < 400 ~ "Low-Intermediate BAL Amylase", amylase >= 400 ~ "High BAL Amylase"))
Time_to_event2 <- Time_to_event2 %>% mutate(Amylase_450=case_when(amylase < 450 ~ "Low-Intermediate BAL Amylase", amylase >= 450 ~ "High BAL Amylase"))
Time_to_event2 <- Time_to_event2 %>% mutate(Amylase_500=case_when(amylase < 500 ~ "Low-Intermediate BAL Amylase", amylase >= 500 ~ "High BAL Amylase"))

#Create dataframe for patients discharged Alive
Alive <- subset(All_samples, Mortality == "Alive")


#Generate time-to-event curve for time to extubation for Alive
fit <- survfit(Surv(intubation_duration) ~ group, data=Alive)
print(fit)
p <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE, risk.table = FALSE, risk.table.col = "strata", linetype = "strata", surv.median.line = "hv", ggtheme = theme_bw(), palette = c("Dark2"), xlim=c(0,45), font.main = c(14), font.x = c(10), font.y = c(14), font.tickslab = c(12, "plain"), xlab = "Time to Extubation in Days", ylab= "", font.legend=c(12), title="Discharged from Hospital Alive", break.x.by=5, legend.title="BAL Amylase Group", legend.labs=c("High","Low-Intermediate"), legend.position="bottom")


fit <- survfit(Surv(intubation_duration, status) ~ Amylase_250, data=Time_to_event2)
print(fit)
p1 <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE, linetype = "strata", surv.median.line = "hv", ggtheme = theme_bw(), palette = c("Dark2"), xlim=c(0,30), font.main = c(14), font.x = c(10), font.y = c(14), font.tickslab = c(12, "plain"), xlab = "Time to Extubation in Days", ylab= "", font.legend=c(12), title="Amylase Threshold 250 U/L", break.x.by=5, legend.title="BAL Amylase (U/L)", legend.labs=c(" Over 250","Less Than 250")) 

fit <- survfit(Surv(intubation_duration, status) ~ Amylase_400, data=Time_to_event2)
print(fit)
p2 <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE, linetype = "strata", surv.median.line = "hv", ggtheme = theme_bw(), palette = c("Dark2"), xlim=c(0,30), font.main = c(14), font.x = c(10), font.y = c(14), font.tickslab = c(12, "plain"), xlab = "Time to Extubation in Days", ylab= "", font.legend=c(12), title="Amylase Threshold 400 U/L", break.x.by=5, legend.title="BAL Amylase (U/L)", legend.labs=c(" Over 400","Less Than 400")) 

fit <- survfit(Surv(intubation_duration, status) ~ Amylase_500, data=Time_to_event2)
print(fit)
p3 <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE,  linetype = "strata", surv.median.line = "hv", ggtheme = theme_bw(), palette = c("Dark2"), xlim=c(0,30), font.main = c(14), font.x = c(10), font.y = c(14), font.tickslab = c(12), xlab = "Time to Extubation in Days", ylab= "", font.legend=c(12), title="Amylase Threshold 500 U/L", break.x.by=5, legend.title="BAL Amylase (U/L)", legend.labs=c("Over 500","Less Than 500")) 

splots <- list()
splots[[1]] <- p
splots[[2]] <- p2
splots[[3]] <- p1
splots[[4]] <- p3

arrange_ggsurvplots(splots, print = TRUE, nrow = 2, ncol = 2)
```

## Find log rank p value for KM curve for patients discharged alive
```{r echo=FALSE}
fit <- survfit(Surv(intubation_duration) ~ group, data=Alive)
print(fit)
surv_pvalue(fit, combine = TRUE, get_coord = TRUE)

```



## Adjust p values for multiple comparisons
```{r, echo=FALSE}
#Create a table with label and p values
p_value_table <- data.frame(Name=c("High-Low Amylase", "Amylase_250", "Amylase_400", "Amylase_500"), p_values=c(0.004, 0.059, 0.0064, 0.0011))
                            
#Enter each p value from the above statistical tests
p_values=c(0.004, 0.059, 0.0064, 0.0011)
p_adjusted <- p.adjust(p_values, method = "bonferroni")
p_value_table$p_adjusted <- p_adjusted <- p.adjust(p_values, method = "bonferroni")
print(p_value_table)
```
